\startcomponent intro
\starttitle[title=Introduction]
Proofs are at the foundations of mathematics.
Viewed through the lens of theoretical computer science, verifying the correctness of a mathematical proof is a fundamental computational task.
Indeed, the P versus NP problem, which deals precisely with the complexity of proof verification, is one of the most important open problems in all of mathematics.
\blank[big]
The complexity-theoretic study of proof verification has led to exciting reenvisionings of mathematical proofs.
For example, probabilistically checkable proofs (PCPs) admit local-to-global structure that allows verifying a proof by reading only a minuscule portion of it.
As another example, interactive proofs allow for verification via a conversation between a prover and a verifier, instead of the traditional static sequence of logical statements.
The study of such proof systems has drawn upon deep mathematical tools to derive numerous applications to the theory of computation and beyond.
\blank[big]
In recent years, such probabilistic proofs received much attention due to a new motivation, delegation of computation, which is the emphasis of this summer school.
This paradigm admits ultra-fast protocols that allow one party to check the correctness of the computation performed by another, untrusted, party.
These protocols have even been realized within recently-deployed technology, for example, as part of cryptographic constructions known as succinct non-interactive arguments of knowledge (SNARKs).
\blank[big]
This summer school has provided an introduction to the field of probabilistic proofs and the beautiful mathematics behind it.
Two complementary courses have been offered:
\startitemize[A]
    \item a foundations course, covering the \quotation{classics} of probabilistic proofs. The material includes seminal results that have found a diverse set of applications in theoretical computer science.
    \item a frontiers course, covering contemporary results in probabilistic proofs. The focus of this course is on proof protocols for delegating computations.
\stopitemize
For the first course on Interactive Proofs, the lectures have been taken by prof. Alessandro Chiesa and are available at \hyphenatedurl{https://www.youtube.com/playlist?list=PLGkwtcB-DfpyjJfxPUdwWpg_ygk2OIp9-}, while the worksheets have been discussed with prof. Nick Spooner.
\blank[2*big]
The topics covered have been: Sumcheck Protocol, Interactive Proofs (IP) for PSPACE, Doubly-Efficient IPs, Zero-Knowledge IPs, Interactive Oracle Proofs (IOP), Linear-Size IOPs for Circuits and Machines, and the limitations of both IPs and IOPs
\blank[big]
For the second course on Probabilistic Checkable Proofs, instead, the lectures have been taken by prof. Tom Gur and are available at \hyphenatedurl{\dot}, while the worksheets have been discussed with Marcel Dall'Agnol.
\blank[2*big]
The topics covered have been: \dots
\stoptitle
\stopcomponent